Syntax Is All You Need: A Universal-Language Approach to Mutant Generation

### 2023-03-29 paper/main.tex
clg fixes

This diversity of tools highlights the growth of widespread interest in mutation
testing.  However, it also demonstrates various ways that bringing mutation testing to
typically polyglot modern projects is cumbersome. Each language (or IR, at
minimum) requires its own underlying mutation and testing framework.
\clg{Alex, can something be said about 
  how many LOC the tools we look at are on average, this early?}

@@

%There
%appears to be a significant risk in relying on any mutation tools other
%than the most widely used and well-supported, such as PIT.
A great deal of research effort has been devoted to devising clever
schemes for reducing the computational burden of mutation testing over
the years;\clg{a few references?} in practice, however, the recent adoption of mutation
testing arguably relies more on the availability of vast (cloud) computing
resources than any particular approach from the literature, beyond
heuristic restrictions on the mutants used (e.g., one per line, or
only some siimple operators).   On the other hand, almost no
research has addressed perhaps the most central issue of
mutation testing: in the absence of a working mutation testing
tool for the language in which one is developing software, one cannot
obtain the benefits of mutation testing.  The absence of such tools is
due to the high human cost of developing and maintaining mutation
\clg{OK I don't mind this argument, BUT it kind of seems in conflict with the
  argument above that says there are a lot of tools and mutation testing is
  popular.  I think it might be worth shifting the emphasis of this paragraph to
  be to motivate source-level mutation testing over IR, because that's missing
  so far and would be good to motivate.  Note that it's also an argument made in
  the abstract, which I think is a particularly good summary of the paper.}
testing tools.  Ideally, mutation testing would be available for
\emph{any} newly proposed programming language, as soon as that
language has any community of users at all.  In the present
state-of-the-art, however, it may be years before even rudimentary
tools appear, and they may ``disappear'' shortly there-after.

@@

\clg{I know we're walking a line on UM anonymity but I think we should cite it
  here, just in the third person.}
Using regular expressions to express mutants has many benefits.  While
much more expressive than simple string replacement, regular
expressions are familliar to most developers.  In principle,
development of complex regular expressions for operators can even be
performed by simply providing examples~\cite{bartoli2014automatic}  or
describing the operator in natural
language~\cite{zhong2018generating}.  
\clg{Just as a note (feel free to read and delete): I tried to cut out our
  hypothetical Python program because I read it and was like OK well then why
  didn't you write it? Or show me if it's so short? You know? And I felt like
  the point could be made without it. I think reference to UM here does the
  trick as well really}
These approaches speak to the potential of string manipulation-based approaches
for mutation testing, without the need for program parsing or analysis.

@@

\begin{table}[hbtp]
\centering
\caption{Rule Examples}
\label{tab:rules}
\resizebox{\columnwidth}{!}{%

@@

\clg{OK this is confusing and I don't know how to fix it.  Two problems.  (1) we
  haven't actually introduced a tool yet so talking about something from its
  implementation is confusing.  We do have kind of a hypothetical tool, though
  I've been removing specific references to it, because it wasn't clear enough
  that it was supposed to represent our ACTUAL implementation and it was
  confusing me.  If we'd like it to introduce an actual implementation, then we
  might undo my changes there, but make it clear we're introducing an
  implementation and not a thought experiment.  Problem (2) is that a lot of
  this paper (and even the intro/abstract to date) is about using parser
  combinators for this, not regular expressions.  So talking about using regular
  expressions in our not-introduced implementation when alter we'll be talking
  about parser combinators as a contribution is also confusing.  Beyond
  ``concretize our hypothetical tool into introduicng actual tool, other
  solutions I can think of are (1) forward ref a section of an implementation and expand slightly this
  sentence to be like ``taken directly from our implementation of a regular
  expression approach to syntactic mutation (Section FOO)'', or (2) be more abstract
  about it here, like ``Table X shows examples of universal rules...we discuss
  them further and how such rules are used in our implementation in Section
  FOO.''.  Maybe those are the same I dunno.}
Table~\ref{tab:rules} shows examples of universal, C-like, and
language-specific rules for a few languages, taken directly from our
implementation.  In all cases, the form of a rule is: {\tt regexp
  ==> replacement}, using Python regular expression and replacement syntax.

@@

One obvious concern with any-language source-based mutation is that every time the
program changes, the cost of invalid mutations must be paid again.
However, the locality of mutants and source changes in fact means this
is seldom required.

The {\tt git merge-file} utility takes a base file and two modified
versions of that file, and produces an automatic merge, using the
usual git merge resolution algorithm.    Because a mutant is a (very
small) source change, this means that in practice updating a mutant to
reflect even large changes to a source file is cost-free, and if the
original mutant was valid, the new mutant will usually also be valid.
New mutant generation is only required for new lines of code and
modified lines of code.  To the extent development is incremental,
therefore, mutant generation can also be incremental.

\section{Beyond Regular Expressions with Parser Parser Combinators}
% Better to frame this as "Beyond regex with PPC" than "PPC instead of RE" 
% instead of RE" because RE pattern matching can be implemented by PPCs,
% they are just pattern matching functions that can be restricted to lexical
% RE-style matching too, and, e.g., comby embeds PCRE regex matching support,
% as a parser combinator). In this sense, PPCs _really are_ a better/more
% general/more universal matching engine abstraction at the expressivity level
% where we care about mutating source syntax with CFG properties.

\clg{OK.  I think maybe ??? This section works better as a subsection in the
  previous section, which can build to ``beyond...'' Then this text can/should
  start with a TINY bit more transition from the regular expression thing we've
  discussed so far to motivate comby ``instead'' (even though it's not really
  instead.)  This is more a point for discussion than a unilateral decree, but
  it feels like it flows better as a complete story.  Otherwise it's not super
  clear what the point of Section 2 was, if Section 3 is the real thing we're
  proposing? Is it? On that note, if we do this, I think we can/should move
  ``Incremental Mutation'' down to implementation even if we didn't do it, just
  as a point of discussion like ``oh btw we could do it if we wanted it's
  easy.'' Otherwise it kind of disrupts the flow AND hierararies of operators
  leads REALLY REALLY nicely into the parse parser combinator approach.}
\clg{Either way regardless this needs a better starting transition haha.}
We implement a lightweight parser-driven approach that attunes mutation
operators to language-general syntax-aware transformations. For example, our
operators can ensure syntactically well-formed outputs (e.g., transformations
must preserve balanced parentheses or braces). More generally, operators are
sensitive to coarse-grained context-free grammar properties. These operators go
beyond the expressivity of regular expressions (which are typically limited to
lexical transformations) to generate \emph{varied} and \emph{valid}
programs during mutation testing.

Parser parser combinators~\cite{vanTonderPPC} is a recent approach enabling
syntax-transformation for multiple languages, and (we show) an apt choice for
mutation testing operators. This key idea is to replace code constructs
common to many languages (e.g., multi-line code blocks delineated by braces)
with simple declarative patterns. Using parser combinators for coarse syntax
matching avoids the complexity of integrating heterogeneous rewrite tools for
mutation testing (e.g., one for Java, one for C++) and human effort to
implement language-specific mutation operators (e.g., where the user has to
learn to write operators for each language tool). The mechanization and
expressive properties of parser parser combinators is covered in prior
literature~\cite{vanTonderPPC}; in this paper we demonstrate that their application is
uniquely effective for universal source-based mutation testing. In
particular, we
build mutation operators with \texttt{comby},\footnote{https://comby.dev} the tool that implements parser parser
combinators to declaratively match and rewrite source code syntax.

Comby is especially suited for matching multi-line code blocks and
disambiguating code from comments and strings---constructs that otherwise
confound and complicate regular expression patterns. Comby supports over 40
languages and implements a generic parser for additional languages to recognize
these syntactic constructs. The following is a brief overview of Comby
definitions and syntax for mutation operators implemented in our approach:

@@

Comby provides mechanisms to further define custom match syntax and behavior
beyond these standard patterns.\footnote{https://comby.dev/docs/advanced-usage\#custom-language-definitions} \texttt{Comby} patterns can also embed regular expressions, subsuming their
expressive power to enable regex-based mutation operators intermixed with
context-free syntax properties.\footnote{https://comby.dev/docs/basic-usage\#using-regular-expressions}

\section{Implementation}

We implemented our approach in a mutation testing tool, written in
about 2,200 lines of Python.  The rules defining mutation operators
for a set of languages including C, C++, Python, Java, Lisp, Swift,
Rust, Solidity, Fe, Vyper, JavaScript, and Go, required another 436
lines; this is the total for comby and regexp-based rules, so each
version requires only 218 lines of rule definitions.  The tool makes
use of parser parser combinators via the {\tt comby-python} module.

@@

@@

Both research questions were addressed by generating and running
mutants for a set of four programming languages: C++, Java, Python,
and Rust.  The criteria for language selection was as follows: C++,
Java, and Python are among the most widely used programming languages
at present.  Furthermore, C++, unlike C, can be notoriously hard to
parse, making development of mutation tools that use an AST and
properly consider mutants of, e.g., code inside a template,
difficult.  Java is the language in which mutation testing has been
studied most extensively in recent literature, and PIT is almost
certainly the most popular and widely used mutation tool.  Python long
suffered from a lack of working and maintained mutation tools, despite
a large number of abandoned efforts, though at present it is served by
multiple maintained, AST-based tools.  Finally, Rust is an ``emerging
language'' with few tools targeting it.
\clg{Kinda want a table (or a pointer to tables) with projects per language.
  Definitely want a pointer to at least Table 6, with the tools we compare to.}

@@

Tables~\ref{tab:table_cpp1}-\ref{tab:table_rust1} compare generated
mutants for ``Regex'' and ``Comby'' modes of our tool.  Invalid
mutants are mutants that the tool produces, but which fail to
compile.  The cost of each such mutant is usually small,
since compilation generally fails early for trivial reasons (e.g.,
{\tt break} or {\tt continue} outside a loop), 
but reducing the number of such mutants directly reduces the cost of
initial mutant generation, and generally indicates a better
``understanding'' of the source being mutated.

>>>>

This diversity of tools highlights the growth of widespread interest in mutation
testing.  However, it also demonstrates various ways that bringing mutation testing to
typically polyglot modern projects is cumbersome. Each language (or IR, at
minimum) requires its own underlying mutation and testing framework.
Mutation tools are not trivial to develop; the smallest
single-language tools we examined were nearly 4KLOC, and the mean size
of a mutation testing tool was over 20K non-comment source lines.

@@

%There
%appears to be a significant risk in relying on any mutation tools other
%than the most widely used and well-supported, such as PIT.
A great deal of research effort has been devoted to devising clever
schemes for reducing the computational burden of mutation testing over
the years (e.g., ~\cite{offutt2001mutation,
  untch1993mutation,KaufmanFAKAJ2022}); in practice, however, the recent adoption of mutation
testing arguably relies more on the availability of vast (cloud) computing
resources than any particular approach from the literature, beyond
heuristic restrictions on the mutants used (e.g., one per line, or
only some siimple operators).   On the other hand, almost no
research has addressed perhaps the most central issue of
mutation testing: in the absence of a working mutation testing
tool for the language in which one is developing software, one cannot
obtain the benefits of mutation testing.  The absence of such tools is
due to the high human cost of developing and maintaining mutation
testing tools.  In particular, new languages may have unstable
grammars, and few tools to aid developers in parsing and manipulating
source code.   Ideally, mutation testing would be available for
\emph{any} newly proposed programming language, as soon as that
language has any community of users at all.  In the present
state-of-the-art, however, it may be years before even rudimentary
tools appear, and they may ``disappear'' shortly there-after.  In
fact, the difficulties of parsing code make bytecode or IL-level
mutation a popular recent choice in tool development, despite the advantages of source-based mutation for
users of mutation testing~\cite{CompareSrcBinary}, including the
ability to read and understand mutants (which is often essential if
one aims to add a test to kill a mutant) and to define mutants in
terms of source-level language (or even project/library)-specific
features.  Moreover, even this arguably less desirable bytecode mutation
approach is not available for languages that do not compile to, e.g.,
Java or LLVM bytecode, such as Haskell.

@@

Using regular expressions to express mutants has many benefits, and
has been demonstrated to be possible, with a relatively small code base~\cite{regexpMut}.  While
much more expressive than simple string replacement, regular
expressions are familliar to most developers.  In principle,
development of complex regular expressions for operators can even be
performed by simply providing examples~\cite{bartoli2014automatic}  or
describing the operator in natural
language~\cite{zhong2018generating}.  
These approaches speak to the potential of string manipulation-based approaches
for mutation testing, without the need for program parsing or analysis.

@@

\begin{table}[hbtp]
\centering
\caption{Regular Expression Rule Examples}
\label{tab:rules}
\resizebox{\columnwidth}{!}{%

@@

Table~\ref{tab:rules} shows examples of universal, C-like, and
language-specific rules for a few languages, taken directly from our
implementation of regular-expression based operators (see Section \ref{sec:imp}).  In all cases, the form of a rule is: {\tt regexp
  ==> replacement}, using Python regular expression and replacement
syntax.  While these rules are non-trivial to write and read, we show
below that using parser parser combinators often simplifies and
clarifies such rules.

@@

\subsection{Beyond Regular Expressions with Parser Parser Combinators}
% Better to frame this as "Beyond regex with PPC" than "PPC instead of RE" 
% instead of RE" because RE pattern matching can be implemented by PPCs,
% they are just pattern matching functions that can be restricted to lexical
% RE-style matching too, and, e.g., comby embeds PCRE regex matching support,
% as a parser combinator). In this sense, PPCs _really are_ a better/more
% general/more universal matching engine abstraction at the expressivity level
% where we care about mutating source syntax with CFG properties.

While regular expressions have been shown to be capable of driving mutant generation, we
propose going beyond regular expressions using a lightweight
parser-driven approach that attunes mutation operators to
language-general syntax-aware transformations. Such operators can
perform tasks that go beyond regular transforms and ensure
syntactically well-formed outputs (e.g., transformations must preserve
balanced parentheses or braces). More generally, operators can be made
sensitive to coarse-grained context-free grammar properties. In short,
these
operators go beyond the expressivity of regular expressions (which are
typically limited to lexical transformations) to generate more
\emph{varied} and more \emph{valid} mutants.

Parser parser combinators~\cite{vanTonderPPC} is a recent approach enabling
syntax-transformation for multiple languages, and (we show) an apt choice for
mutation testing operators. This key idea is to replace code constructs
common to many languages (e.g., multi-line code blocks delineated by braces)
with simple declarative patterns. Using parser combinators for coarse syntax
matching avoids the complexity of integrating heterogeneous rewrite tools for
mutation testing (e.g., one for Java, one for C++) and human effort to
implement language-specific mutation operators (e.g., where the user has to
learn to write operators for each language tool). The mechanization and
expressive properties of parser parser combinators is covered in prior
literature~\cite{vanTonderPPC}; in this paper we demonstrate that their application is
uniquely effective for universal source-based mutation testing. In
particular, we
build mutation operators with Comby\footnote{https://comby.dev}, the tool that implements parser parser
combinators to declaratively match and rewrite source code syntax.

Comby is especially suited for matching multi-line code blocks and
disambiguating code from comments and strings---constructs that otherwise
confound and complicate regular expression patterns. Comby supports over 40
languages and implements a generic parser for additional languages to recognize
their syntactic constructs. The following is a brief overview of Comby
definitions and syntax for mutation operators implemented in our approach:

@@

Table~\ref{tab:crules} shows how the rules in Table~\ref{tab:rules}
look, when implemented as Comby patterns.  While some simple rules are
unchanged, others are either more succinct or provide more context as
to applicability (e.g., only less than between valid holes should be
mutated).  This shows that a
further major benefit of using Comby: once a user has learned
the simple concepts above, it is often much easier to read---and
especially \emph{write}---mutation
operator definitions in Comby than using regular expressions.  The
added precision comes not only ``for free'' but in fact yields a
benefit for the ``end-user-programmer'' of mutations.

Comby provides mechanisms to further define custom match syntax and behavior
beyond these standard patterns.\footnote{https://comby.dev/docs/advanced-usage\#custom-language-definitions} Comby patterns can also embed regular expressions, subsuming their
expressive power to enable regex-based mutation operators intermixed with
context-free syntax
properties\footnote{https://comby.dev/docs/basic-usage\#using-regular-expressions}.

\begin{table}[hbtp]
\centering
\caption{Parser Parser Combinator Rule Examples}
\label{tab:crules}
\resizebox{\columnwidth}{!}{%

  \begin{tabular}{|l|l|l|}
    \hline
    Universal & C-Like & Language-Specific \\
    \hline
{\tt :[a]<:[b] ==> :[a]>:[b]} &{\tt  else ==>} & {\tt synchronized ==>} \\ 
{\tt :[a]<:[b] ==> :[a]==:[b]} & {\tt if (:[cond]) ==> if (!(:[cond]))} &
                                                                 {\tt
                                                                          :[ s]:[expr]:[lf~[\textbackslash n]] ==> :[s]pass:[lf]}\\
{\tt :[a]<:[b] ==> :[a]<=:[b]} & {\tt if (:[cond]) ==> if (1==1)} & {\tt any\_of ==> all\_of}\\
\hline

\section{Implementation}
\label{sec:imp}

We implemented our approach in a mutation testing tool, written in
about 2,200 lines of Python.  The rules defining mutation operators
for a set of languages including C, C++, Python, Java, Lisp, Swift,
Rust, Solidity, Fe, Vyper, JavaScript, and Go, required another 436
lines; this is the total for Comby and regexp-based rules, so each
version requires only 218 lines of rule definitions.  The tool makes
use of parser parser combinators via the {\tt comby-python} module.

@@

One obvious concern with any-language source-based mutation is that every time the
program changes, the cost of invalid mutants must be paid again.
However, the locality of mutants and source changes in fact means this
is seldom required.

The {\tt git merge-file} utility takes a base file and two modified
versions of that file, and produces an automatic merge, using the
usual git merge resolution algorithm.    Because a mutant is a (very
small) source change, this means that in practice updating a mutant to
reflect even large changes to a source file is cost-free, and if the
original mutant was valid, the new mutant will usually also be valid.
New mutant generation is only required for new lines of code and
modified lines of code.  To the extent development is incremental,
therefore, mutant generation can also be incremental.  Our
implementation allows the use of {\tt git merge-file} to control the
modification of most mutants, and uses our mechanism for
selective mutant generation based on coverage to force mutation where
there is a conflict in a ``mutant patch.''

@@

Both research questions were addressed by generating and running
mutants for a set of four programming languages: C++, Java, Python,
and Rust.  The criteria for language selection was as follows: C++,
Java, and Python are among the most widely used programming languages
at present.  Furthermore, C++, unlike C, can be notoriously hard to
parse, making development of mutation tools that use an AST and
properly consider mutants of, e.g., code inside a template,
difficult.  Java is the language in which mutation testing has been
studied most extensively in recent literature, and PIT is almost
certainly the most popular and widely used mutation tool.  Python long
suffered from a lack of working and maintained mutation tools, despite
a large number of abandoned efforts, though at present it is served by
multiple maintained, AST-based tools.  Finally, Rust is an ``emerging
language'' with few tools targeting it.  Table~\ref{tab:mutationtools}
gives an overview of tools we identified for each language as working tools
capable of single-file mutant generation.

@@

Tables~\ref{tab:table_cpp1}-\ref{tab:table_rust1} compare generated
mutants for Regex and Comby modes of our tool.  Invalid
mutants are mutants that the tool produces, but which fail to
compile.  The cost of each such mutant is usually small,
since compilation generally fails early for trivial reasons (e.g.,
{\tt break} or {\tt continue} outside a loop), 
but reducing the number of such mutants directly reduces the cost of
initial mutant generation, and generally indicates a better
``understanding'' of the source being mutated.
